Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-09 paper.qmd
revised discussion and supp materials and change round brackets

- Make sure langevitour is explained before using it
- Too much code needs to be streamlined
- Remove all the extraneous commented text
- Ensure that the SCdeed paper is mentioned in main part of this paper

@@

With our goal being to make a representation of this \gD{} layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. We recommend using a hexagon grid because it better reflects the data distribution and has less artifacts than a rectangular grid. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-two-curvy, and documented below.

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighbouring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connectd centroids. The edges preserve the neighborhood information when the model is lifted into \pD{}.

@@

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This seingle projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbours in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data. <!-- likely that the points are organised by the global principal components. So the layout of the model in \fD{} doesn't reflect the dimension of each cluster. This can be happen because PaCMAP balances three components (near, mid-range, and far) rather than focusing solely on near neighbors. Mid-range neighbors ensure that transitions between clusters or groups are preserved and less concern with local preservation.-->

@@

Single-cell data refers to the characteristics of individual cells within a population (@haque2017). These data also provides insights at the level of individual cells, enabling deeper understanding of gene expression.

@@

We have developed an algorithm to assess the NLDR method and (hyper-)parameter choice(s) and choose the most accurate representation of the structure(s) present in the \pD{} data. Starting from \gD{} layout, the fitted model, as represented by the positions of points in \kD{}, and turn it into a \pD{} wireframe to overlay on the data, viewing it with a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

@@

Our algorithm can also be useful in practice to decide on a reasonable NLDR layout that accurately represents the data, as described with a single-cell gene expression data, **PBMC3k**. The published results of UMAP with default (hyper-)parameter setting by @chen2024 were evaluated. The data showed three close clusters. But the published layout showed three well-separated clusters. Therefore, by changing the NLDR method to tSNE with default (hyper-)parameters, we found a better layout, which is more representative the data. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the \gD{} layout, demonstrating the ability of our model in uncovering hidden data characteristics. Also, as explained in Appendix, with recently introduced approach, scDEED also proved the layout suggested by our layout is better than the published one.

@@

Other tour software instead of langevitour.

@@

Appendix: The appendix includes more details about the hexagonal binning algorithm and scDEED method (appendix.pdf, Portable Document Format file).

R package `quollr`: The R package `quollr` containing codes to fit, and visualize the model. (need to add quollr .zip file, GNU zipped tar file)

@@

These `R` packages were used for the work: `tidyverse` (@hadley2019), `Rtsne` (@jesse2015), `umap` (@tomasz2023), `patchwork` (@thomas2024), `colorspace` (@achim2020), `langevitour` (@harisson2024), `conflicted` (@hadley2023), `reticulate` (@kevin2024), `kableExtra` (@hao2024). These `python` packages were used for the work: `trimap` (@amid2019) and `pacmap` (@yingfan2021). The article was created with `R` packages `quarto` (@jjallaire2024). The project's GitHub repository ([https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm)) contains all materials required to reproduce this article.

>>>>

- Make sure langevitour is explained before using it
- Too much code needs to be streamlined
- Remove all the extraneous commented text
- Check how \kD{} and \gD{} are used, and make sure it's sensible

@@

With our goal being to make a representation of this \gD{} layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbor information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. We recommend using a hexagon grid because it better reflects the data distribution and has less artifacts than a rectangular grid. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-two-curvy, and documented below.

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighboring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connectd centroids. The edges preserve the neighborhood information when the model is lifted into \pD{}.

@@

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This seingle projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbors in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data. <!-- likely that the points are organised by the global principal components. So the layout of the model in \fD{} doesn't reflect the dimension of each cluster. This can be happen because PaCMAP balances three components (near, mid-range, and far) rather than focusing solely on near neighbors. Mid-range neighbors ensure that transitions between clusters or groups are preserved and less concern with local preservation.-->

@@

Single-cell data refers to the characteristics of individual cells within a population [@haque2017]. These data also provides insights at the level of individual cells, enabling deeper understanding of gene expression.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)parameter choice(s). It depends on conceptualising the \gD{} layout as a model, creating a wireframe representation of it that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

Two examples illustrating usage are provided: the PBMC3k data where the NLDR is summarising clustering in \pD{} and hand-written digits illustrating how NLDR represents an intrinsically lower dimensional nonlinear manifold. We demonstrated that the published layout of the PBMC3k is inaccurate, because it grossly exaggerates separation between clusters, and even suggests separation when there is none. This is common when layouts are chosen subjectively -- often a preference for the "prettiest". Our approach provides a way to objectively choose the layout and hopefully avoids the use of misleading layouts in the future. In the hand-written digits we illustrate how our model fit statistics show that a flat disc layout is superior to the curved shaped layouts, and how to identify oddly written "1"s using the residuals of the fitted model.

Additional exploration of metrics to summarise the fit could be a new direction for the work. The difficulty is capturing nonlinear fits, for which Euclidean distance can be sub-optimal. We have used a very simple approach based on clustering methods, Euclidean distances to nearest centroid, which can approximate nonlinear patterns. Other cluster metrics would be natural choices to explore.

This new method also reveals some interesting curiosities about NLDR procedures. The fitted model appears as a "pancake" in some data where clusters are regularly shaped and high-dimensional, for some methods but not others, which is odd. One can imagine that if algorithms are initiated using principal components then some ordering of points along the major axes might generate this pattern. Alternatively, if local distances dominate the algorithm then is might be possible to see this pattern with well-separated regular clusters. We also demonstrated that there is a tendency for NLDR algorithms to be confused by different density in the data space, and some patterns in the layout are due to density differences rather than nonlinear associations between variables.

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used asa the basis for a wireframe of the fitted model. @barber1996 and the software @cxhull have algorithms for convex hulls, which \pD{} which serve as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.

The new methodology is accompanied by an R package called quollr, so that it is readily usable and broadly accessible. The package has methods to fit the model, compute diagnostics and also visualise the results, with interactivity. We have primarily used the langevitour software [@harisson2024] to view the model in the data space, but other tour software such as tourr [@tourr] and [@detourr] could be also used.

<!-- and choose the most accurate representation of the structure(s) present in the \pD{} data. Starting from \gD{} layout, the fitted model, as represented by the positions of points in \kD{}, and turn it into a \pD{} wireframe to overlay on the data, viewing it with a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

@@

Our algorithm can also be useful in practice to decide on a reasonable NLDR layout that accurately represents the data, as described with a single-cell gene expression data, **PBMC3k**. The published results of UMAP with default (hyper-)parameter setting by @chen2024 were evaluated. The data showed three close clusters. But the published layout showed three well-separated clusters. Therefore, by changing the NLDR method to tSNE with default (hyper-)parameters, we found a better layout, which is more representative the data. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the \gD{} layout, demonstrating the ability of our model in uncovering hidden data characteristics. Also, as explained in Appendix, with recently introduced approach, scDEED [@xia2023] also proved the layout suggested by our layout is better than the published one.

@@

Other tour software instead of langevitour. 
-->

@@

The GitHub repository for the paper can be found at https://github.com/JayaniLakshika/paper-nldr-vis-algorithm. The Appendix includes more details about the hexagonal binning algorithm and a comparison to the results of the newly reported scDEED [@xia2023] statistic.

The R package `quollr`, available on CRAN and at https://jayanilakshika.github.io/quollr/, provides software accompaying this paper to fit the wireframe model representation, compute diagnostics, visualize the model in the data with langevitour and link multiple plots interactively.

@@

These `R` packages were used for the work: `tidyverse` [@hadley2019], `Rtsne` [@jesse2015], `umap` [@tomasz2023], `patchwork` [@thomas2024], `colorspace` [@achim2020], `langevitour` [@harisson2024], `conflicted` [@hadley2023], `reticulate` [@kevin2024], `kableExtra` [@hao2024]. These `python` packages were used for the work: `trimap` [@amid2019] and `pacmap` [@yingfan2021]. The article was created with `R` packages `quarto` [@jjallaire2024].
